### 2015-06-09 main.tex
Integrated Dan's newest comments.

%                 Doubly-Stopped Binomial Distribution

%                        Kane and Zelterman

\typeout{}\typeout{}\typeout{}\typeout{}
\typeout{ Two-Stage, Phase II Designs }
\typeout{}\typeout{}
\typeout{ Michael Kane \& Dan Zelterman }
\typeout{}\typeout{}\typeout{}\typeout{}

@@

%\font\fiverm=cmr5                     % Font definition for PicTex
%\input{prepic.tex}                    % Input three PicTeX files
%\input{pictex.tex}
%\input{postpic.tex}

@@

%  These commands allow exercises listed as \item[]'s to be listed in
%  the .aux file and referenced in the text.

@@

In the classic Simon 2-stage Optimal Design, Phase II clinical trial [\ref{Simon 1989}] a small number of patients are enrolled onto a therapeutic protocol for an initial stage and assess either success according to a clinically useful criteria or else as treatment failure. Failure might include adverse reaction to the drug or withdrawal due to toxicity. We use a large number of these early successes as evidence for whether to additional patients for a second stage. A statistically small number of successes lead us to terminate the trial.

Consider the Phase 2 clinical trial that investigates the efficacy of iniparib (BSI-201/SAR240550) in patients with breast cancer gene-associated (BRCA) ovarian cancer [\ref{Sanofi 2013}]. In the first stage 12 patients were treated with iniparib (BSI-201/SAR240550). If at least two of those patients respond, as defined by Response Evaluation Criteria in Solid Tumor (RECIST), then the trial proceeds to the next stage where additional patients were treated.

If, in the first stage, two or more are successful then we are lead to believe that the therapy is useful and will proceed to the second stage. Otherwise, the trial will terminate. Our goal is to know the distribution of the sample size of the design for planning purposes.
If all 12 patients are enrolled at once, as in the classic design, then the sample size is 12. However, in most clinical trials, the patients are enrolled sequentially, often with one patient's outcome realized before the next one enters the trial. In the present example, observing two successful patients allows us to proceed to the next stage. Similarly 11 observed treatment failures also ends the stage. This sampling mechanism, in which the experiment ends as soon as any of the endpoints is reached, is call {\em curtailed sampling}. Under curtailed sampling the range of the sample size is between two and 11.

@@

Let $\,B_1, B_2, \ldots \,$ denote a sequence of independent, identically distributed, Bernoulli random variables with 
$\mathbb{P}\{B_i=1\}=p$, for probability parameter $p \in \{0\leq p\leq 1\}$. Let $s$ and $t$ be positive integers. Define the Stopped Negative Binomial (SNB) random variable $Y$ as the smallest integer value such that $\,\{B_1, \ldots , B_Y\}\,$ contains either $\,s\,$ successes or else $\,t\,$ failures. The distribution of $\,Y\,$ has support on integer values in the range 
\begin{equation*}                                     %   (3)
     \min(s,t) \leq \; Y \;\leq s+t-1  \label{range.y.eq}
\end{equation*}
and it is distributed as
\begin{equation} \label{eqn:pmf}
\mathbb{P}\{Y=k\} = S(k, p, s)  \{s \leq k \leq s+t-1\} + T(k, p, t) \{t \leq k \leq s+t-1 \} 
\end{equation}
where
\begin{equation} \label{eqn:N}
S(k, p, s) = {k-1 \choose s-1} p^s (1-p)^{k-s} 
\end{equation}
is the negative binomial probability of $s$ successes and
\begin{equation} \label{eqn:R}
T(k, p, t) = {k-1 \choose k-t} p^{k-t} (1-p)^t
\end{equation}
is the negative binomial probability of $t$ failures.

@@

To show that (\ref{eqn:pmf}) is the distribution of $Y$, consider the process $\mathbf{X} = \{X(k) : k \in \mathbb{Z}_{\ge 0} \}$ with $X(0)=0$ and
\begin{equation*} \label{eqn:proc}
X_{k+1} = X_k + B_{k+1} \{ s \leq X_k \leq k-t \}
\end{equation*}
where $B_{k+1}$ is once again distributed as Bernoulli($p$) for all $k$ in the domain of $\mathbf{X}$. The process can be conceptualized as a series of coin flips that stops when either $s$ heads or $t$ tails are reached. At each step a coin is flipped. If it is heads, the process advances one diagonally in the positive horizontal and vertical direction. Otherwise, it advances in the positive horizontal direction only. The process stops, $X_k$'s value does not change, after either $X_k = s$ or $X_k = k-t$. One example is given in Fig. \ref{fig:kane_viz}.

@@

The probability that a given realization of $\mathbf{X}$ reaches $s$ at time $k$ is  the probability that, at time $k-1$ there are $s-1$ successful coin flips and $k-s$ unsuccessful coin flips times the probability of a successful coin flip at time $k$ (\ref{eqn:N}). The probability $X_k = k-t$ is the probability that, at time $k-1$ there are $k-t$ successful coin flips and $t-1$ unsuccessful coin flips times the probability of an unsuccessful coin flip at time $k$ (\ref{eqn:R}).

Next, we need to show that the sum
\begin{equation} \label{eqn:sum_proof}
R = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s (1-p)^{k-s} + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} (1-p)^t
\end{equation}
is equal to one.
First, substitute $i=k-s$ in the first summation and
$j=k-t$ in the second. Then (\ref{eqn:sum_proof}) becomes:
\begin{equation*}
R = \sum_{i=0}^{t-1} {i+s-1 \choose s-1} p^s (1-p)^i +
\sum_{j=0}^{s-1} {j+t-1 \choose j} p^j (1-p)^t.
\end{equation*}
Write   
\begin{equation*}
{j+s-1 \choose s-1} = {j+s-1 \choose j}
\end{equation*}
and so $R$ can be written
as the cumulative distribution function of two 
negative binomial distributions
\begin{equation} \label{eqn:transformed_sum}
R = \sum_{i=0}^{t-1} {i+s-1 \choose i} p^s (1-p)^i +
\sum_{j=0}^{s-1} {j+t-1 \choose j} p^j (1-p)^t.
\end{equation}
Let $I_p(s, t)$ be the {\em regularized incomplete beta function}, which is also one minus the c.d.f. of the negative
binomial distribution, and $I_p(s, t) = I_{1-p}(t, s)$.
Then the result can be proven as:
\begin{align*}
R = \sum_{i=0}^{t-1} &{i+s-1 \choose i} p^s (1-p)^i +
\sum_{j=0}^{s-1}  {j+t-1 \choose j} p^j  (1-p)^t \\
   &= 1-I_p(s, t) + 1 - I_{1-p}(t, s) \\
   &= 1. 
\end{align*}
\end{proof}
From (\ref{eqn:pmf}) one can see that the mass at any point on the support of the distribution may come from either $X_k = s$, captured in the function $S$, or $X_k = k-t$, captured in the function $T$. The support of $S$ is defined by $s \leq S \leq s+t+1$ and the support of $T$ is defined by $t \leq T \leq s+t+1$.

@@

The probability mass function of $\,Y\,$ has a variety of shapes for different choices of the parameters $(s,\, t,\,p)$.
These shapes are illustrated in Fig.~\ref{shapes.fig}.
The SNB is related to the negative binomial distribution. Specifically, if $t$ is infinitely large then the $Y-s$ has a negative binomial distribution with
\begin{equation*}                                    %   (1)
   \mathbb{P}\{Y=s+j \}        \label{nb1.eq}
          = {{s+j-1}\choose{s-1}} p^s (1-p)^j
\end{equation*}
for $\,j=0, 1,\ldots\,$.
Similarly, if $s$ is infinitely large then $\,Y - t\,$ has a negative binomial distribution with
\begin{equation*}                                    %   (2)
    \mathbb{P}\{Y=t+j \}        \label{nb2.eq} 
               = {{t+j-1}\choose{t-1}} p^j(1-p)^t.
\end{equation*}

>>>>

%                 Doubly-Stopped Binomial Distribution
%                        Kane and Zelterman
\typeout{}\typeout{}\typeout{}\typeout{}
\typeout{ Two-Stage, Phase II Designs }
\typeout{}\typeout{}
\typeout{ Michael Kane \& Dan Zelterman }
\typeout{}\typeout{}\typeout{}\typeout{}

@@

@@

@@

In the Simon 2-stage Phase II clinical trial [\ref{Simon 1989}], a small 
number of patients are enrolled onto a therapeutic 
protocol for an initial stage. Fore each patient, we assess either success 
or failure according to a 
clinically useful criteria. Success might include tumor response to treatment
or perhaps that the response is maintained for a minimum specified duration.
In addition to failing to achieve the primary endpoint, treatment failure 
might include adverse reaction to the drug or withdrawal due to toxicity. We 
use a large number of these early successes as evidence for whether to enroll
additional patients for a second stage. A statistically small number of 
successes in Stage I would lead us to terminate the trial without entering 
Stage II.

Consider the Phase II clinical trial that investigates the efficacy of 
iniparib in patients with breast cancer gene-associated (BRCA) ovarian 
cancer [\ref{Sanofi 2013}]. In the first stage 12 patients were treated with 
iniparib (BSI-201/SAR240550). The endpoint was defined by Response Evaluation 
Criteria in Solid Tumor (RECIST). If at least two Stage I patients succeeded,
then the trial would proceed to the next stage where additional patients were 
treated. If fewer than two of the 12 Stage I patients is successful then 
the trial would terminate.

Our goal is to know the distribution of the sample size of the design for 
planning purposes.  If all 12 patients are enrolled at once, as in the classic 
design, then the sample size is 12. However, in most clinical trials, the 
patients are enrolled sequentially, often with one patient's outcome realized 
before the next one enters the trial. In the present example, observing two 
successful patients allows us to proceed to the next stage. Similarly 11 
observed treatment failures also ends the stage. This sampling mechanism, in 
which the experiment ends as soon as any of the endpoints is reached, is 
call {\em curtailed sampling}. Under curtailed sampling the range of the 
sample size is between two and 12.

@@

Let $\,B_1, B_2, \ldots \,$ denote a sequence of independent, identically 
distributed, Bernoulli random variables with $\mathbb{P}\{B_i=1\}=p$, for 
probability parameter $0\leq p \leq 1$. In the clinical trial setting 
$\,B_i = 0$ corresponds to a successful enrollment. 
Let $s$ and $t$ be positive integers. 
Define the stopped negative binomial (SNB) random variable $Y$ as the smallest 
integer value such that $\,\{B_1, \ldots , B_Y\}\,$ contains {\em either} 
$\,s\,$ successes or else $\,t\,$ failures. The distribution of $\,Y\,$ has 
support on integer values in the range 
\begin{equation*}                                     %   (3)
     \min(s,t) \leq \; Y \;\leq s+t-1  \label{range.y.eq}
\end{equation*}
and it is distributed as
\begin{equation} \label{eqn:pmf}
\mathbb{P}\{Y=k\} = S(k, p, s)  \{s \leq k\} + T(k, p, t) \{t \leq k \} 
\end{equation}
where
\begin{equation} \label{eqn:N}
S(k, p, s) = {k-1 \choose s-1} p^s (1-p)^{k-s} 
\end{equation}
is the negative binomial probability of $s$ successes and
\begin{equation} \label{eqn:R}
T(k, p, t) = {k-1 \choose k-t} p^{k-t} (1-p)^t
\end{equation}
is the negative binomial probability of $t$ failures.

@@

To illustrate that (\ref{eqn:pmf}) is the distribution of $Y$, consider the 
process $\mathbf{X} = \{X(k) : k \in \mathbb{Z}_{\ge 0} \}$ with $X(0)=0$ and
\begin{equation*} \label{eqn:proc}
X_{k+1} = X_k + B_{k+1} \{ k-t < X_k < s \}
\end{equation*}
where $B_{k+1}$ is once again distributed as Bernoulli($p$) for all $k$ in the domain of $\mathbf{X}$. The process can be conceptualized as a series of coin flips that stops when either $s$ heads or $t$ tails are reached. At each step a coin is flipped. If it is heads, the process advances one diagonally in the positive horizontal and vertical direction. Otherwise, it advances in the positive horizontal direction only. The process stops, $X_k$'s value does not change, after either $X_k = s$ or $X_k = k-t$. One example is given in Fig. \ref{fig:kane_viz}.

@@

The probability that a given realization of $\mathbf{X}$ reaches $s$ at 
time $k$ is  the probability that, at time $k-1$ there are $s-1$ successful 
coin flips and $k-s$ unsuccessful coin flips times the probability of a 
successful coin flip at time $k$ (\ref{eqn:N}). The probability $X_k = k-t$ 
is the probability that, at time $k-1$ there are $k-t$ successful coin flips 
and $t-1$ unsuccessful coin flips times the probability of an unsuccessful 
coin flip at time $k$ (\ref{eqn:R}).  Next, we need to show that the sum
\begin{equation} \label{eqn:sum_proof}
R = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s (1-p)^{k-s} + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} (1-p)^t
\end{equation}
is equal to one.
First, substitute $i=k-s$ in the first summation and
$j=k-t$ in the second. 
\begin{equation*}
{j+s-1 \choose s-1} = {j+s-1 \choose j}
\end{equation*}
Then %(\ref{eqn:sum_proof}) becomes:
%\begin{equation*}
%R = \sum_{i=0}^{t-1} {i+s-1 \choose s-1} p^s (1-p)^i +
%\sum_{j=0}^{s-1} {j+t-1 \choose j} p^j (1-p)^t.
%\end{equation*}
$R$ can be written
as the cumulative distribution function of two 
negative binomial distributions
\begin{equation} \label{eqn:transformed_sum}
R = \sum_{i=0}^{t-1} {i+s-1 \choose i} p^s (1-p)^i \; + \;
\sum_{j=0}^{s-1} {j+t-1 \choose j} p^j (1-p)^t.
\end{equation}

Let $I_p(s, t)$ be the {\em regularized incomplete beta function}, which is 
also one minus the c.d.f. of the negative binomial distribution. This
function satisfies $I_p(s, t) = I_{1-p}(t, s)$.  Then 
\begin{align*}
R = \sum_{i=0}^{t-1} &{i+s-1 \choose i} p^s (1-p)^i +
\sum_{j=0}^{s-1}  {j+t-1 \choose j} p^j  (1-p)^t \\
   &= 1-I_p(s, t) + 1 - I_{1-p}(t, s) \\
   &= 1. 
\end{align*}
\end{proof}
Proving that \label{eqn:pmf} is a valid probability mass function.

From (\ref{eqn:pmf}) one can see that the mass at any point on the support of the distribution may come from either $X_k = s$, captured in the function $S$, or $X_k = k-t$, captured in the function $T$. The support of $S$ is defined by $s \leq S \leq s+t+1$ and the support of $T$ is defined by $t \leq T \leq s+t+1$.

@@

The probability mass function of $\,Y\,$ has a variety of shapes for different choices of the parameters $(s,\, t,\,p)$.
These shapes are illustrated in Fig.~\ref{shapes.fig}.
The SNB is related to the negative binomial distribution. Specifically, if 
$t$ is infinitely large then the $Y-s$ has a negative binomial distribution 
with
\begin{equation*}                                    %   (1)
   \mathbb{P}\{Y=s+j \}        \label{nb1.eq}
          = {{s+j-1}\choose{s-1}} p^s (1-p)^j
\end{equation*}
for $\,j=0, 1,\ldots\,$. A similar statement can be made when $s$ is large
and $t$ is small.
